Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-08 manuscript.rmd
1. coefficient valeus given for models of A & B 2. tiff instead of jpeg

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapour pressure deficit, matric potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees' mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
\
Keywords: belowground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies

@@

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information. 
\
In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere[xx add references].

@@

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D). Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]")}
knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
```

@@

\begin{equation}
A = a_3WSG^2 + a_2B + a_1B^2 + a_0 + error
(\#eq:a-mod)
\end{equation}

\begin{equation}
B = b_3WSG^2 + b_2WSG.LMA + b_1LMA^2 + b_0 + error
(\#eq:b-mod)
\end{equation}

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-vuln-curves)).

```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
```

@@

## Mortality Analyses and Extreme Drought Experience
\
We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
\
All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).

@@

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.jpeg"))
```

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and jittered vertically around the named depths (ticks) for clarity. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors. Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the cenus intervals. Extreme drought periods from multiple years may be overlaid within a panel. Wet season is shown in gray as beginning with DOY."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

@@

```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

@@

```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
```

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B). Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```

>>>>

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapour pressure deficit, matric potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees' mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

\
Keywords: belowground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies

@@

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.

\
In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; Leigh.1982; @Gentry.1990].

@@

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D). Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]xx")}
knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
```

@@

\begin{equation}
A = `r acf[1]` `r acf[2]`WSG `r acf[3]`WSG^2 + `r acf[4]`WSG.B + `r acf[5]`B `r acf.6`B^2 + error
(\#eq:a-mod)
\end{equation}

\begin{equation}
B = `r bcf[1]` `r bcf[2]`B + `r bcf[3]`B^2 + `r bcf[4]`B.LMA + `r bcf[5]`LMA `r bcf[6]`LMA^2 + error
(\#eq:b-mod)
\end{equation}

$WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-vuln-curves)).

```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.tiff"))
```

@@

## Mortality Analyses and Extreme Drought Experience
\
We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against species ($Mort_{mean}$). To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
\
All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).

@@

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.tiff"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.tiff"))
```

@@

\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.tiff"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and jittered vertically around the named depths (ticks) for clarity. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.tiff"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors. Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the cenus intervals. Extreme drought periods from multiple years may be overlaid within a panel. Wet season is shown in gray as beginning with DOY."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.tiff"))
```

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

@@

```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.tiff"))
```

@@

```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.tiff"))
```

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B). Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.tiff", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.tiff", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```
